---
title: "RBaM"
subtitle: "An R user interface to BaM"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,results='hide',
                      fig.path = "man/readme/README-")
```

# Introduction

BaM (Bayesian Modeling) is a framework to estimate a model using Bayesian inference. The R package `RBaM` is built as an R User Interface to the [computational BaM engine](https://github.com/BaM-tools/BaM). It defines classes (objects' properties and methods) for the various building blocks of a BaM case study. Its typical usage is as follows:

1. Assemble the dataset of input/ouput variables.
2. Define the model and set prior distributions for its parameters.
3. Perform Bayesian-MCMC inference.
4. Perform predictions.

```{r message=FALSE}
# devtools::install_github('BaM-tools/RBaM') # First use: install the package from GitHub
library(RBaM) # Load package
```

The catalogue of distributions and models that are available in RBaM can be accessed as follows:

```{r results='markup'}
getCatalogue()
```

# A simple example using the BaRatin rating curve model






