---
title: "TextFile Model"
subtitle: "Defining a simple model by typing formulas"
author: "Ben Renard"
date: "28/07/2020"
output: 
  html_document: 
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,results='hide')
```
# Introduction and aims
This vignette demonstrates how to define a new model by typing its constitutive formulas, and how to use `BaM` to calibrate and use it.

# Description of the model
The model assumes that the population of a given species at time $t$ is equal to:
$$P(t)=\frac{K}{1+\frac{K-P_0}{P_0}*exp(-r*T(t)*t)}$$
Where $P_0>0$ is the initial population, $K>P_0$ is the population upper limit (linked to environmental constraints), $T(t)>0$ is the temperature at time $t$ (assumed to be always positive here) and $r$ is the growth-to-temperature ratio (i.e. the growth rate is equal to $r*T$ - the warmer the temperature, the larger the population growth).

Assume that the same initial population $P_0$ of the same species (hence having the same $r$) is placed in two different environments (hence different upper limits $K_1$ and $K_2$ and different temperatures times series $T_1$ and $T_2$). This lead to a model with two output variables ($P_1$ and $P_2$), three input variables ($t$,$T_1$ and $T_2$) and four parameters ($P_0$,$r$, $K_1$ and $K_2$) defined by the formulas:
$$P_1(t)=\frac{K_1}{1+\frac{K_1-P_0}{P_0}*exp(-r*T_1(t)*t)}$$
$$P_2(t)=\frac{K_2}{1+\frac{K_2-P_0}{P_0}*exp(-r*T_2(t)*t)}$$

# Implementation in BaM
Let's start by loading the `RBaM` package and define the workspace where config and result files will be written.

```{r}
library(RBaM)
workspace=file.path(tempdir(),'BaM_workspace')
```

The model is defined by two formulas, so let's type them as simple strings.

```{r}
f1='K1/(1+((K1-P0)/P0)*exp(-r*T1*t))'
f2='K2/(1+((K2-P0)/P0)*exp(-r*T2*t))'
```

The next step is to define `BaM` `model` object. The main point to be careful with is to make sure that the parameter names are the same as the one used in the formulas. The name of the inputs and the formulas themselves are passed through the `xtraModelInfo` object.

```{r}
# parameters
p1=parameter(name='P0',init=100,prior.dist='Gaussian',prior.par=c(100,10))
p2=parameter(name='r',init=0.001,prior.dist='LogNormal',prior.par=c(-7,1))
p3=parameter(name='K1',init=10000,prior.dist='LogNormal',prior.par=c(9,1))
p4=parameter(name='K2',init=10000,prior.dist='LogNormal',prior.par=c(9,1))
# use xtraModelInfo to pass the names of the inputs and the formulas
xtra=xtraModelInfo(object=list(inputs=c('t','T1','T2'),formulas=c(f1,f2)))
# model
mod=model(ID='TextFile',nX=3,nY=2,par=list(p1,p2,p3,p4),xtra=xtra)
```

# Model calibration

Let's first define calibration data (synthetic data are used here). The data frame `twoPopulations` comes with the `RBaM` package. 

```{r}
# Input data. WARNING: make sure column order is consistent with inputs declared in xtraModelInfo
X=twoPopulations[,1:3] 
# Output data. WARNING: make sure column order is consistent with the order of formulas
Y=twoPopulations[,4:5]
# dataset object
data=dataset(X=X,Y=Y,data.dir=workspace)
```

Model calibration can now be performed by calling the function `BaM`.

```{r,results='asis'}
BaM(workspace=workspace,mod=mod,data=data)
```

Visualizing MCMC simulations allows verifying that MCMC convergence is satisfying and allows exploring estimated parameters.

```{r,results='asis'}
mcmc=readMCMC(file=file.path(workspace,'Results_Cooking.txt'))
g=tracePlot(mcmc)
gridExtra::grid.arrange(grobs=g,ncol=3)
g=densityPlot(mcmc)
gridExtra::grid.arrange(grobs=g,ncol=3)
```

Observed vs. modeled population sizes can easily be compared thanks to the `Residuals` result file.

```{r,results='asis'}
res=read.table(file=file.path(workspace,'Results_Residuals.txt'),header=TRUE)
res[res==-9999] <- NA
library(ggplot2)
g=ggplot(res,aes(x=X1_obs))+theme_bw()+xlab('Time')+ylab('Population size')
g=g+geom_line(aes(y=Y1_sim),color='blue')+geom_point(aes(y=Y1_obs),color='blue')
g=g+geom_line(aes(y=Y2_sim),color='red')+geom_point(aes(y=Y2_obs),color='red')
print(g)
```

Finally, it is possible to use the calibrated model to make predictions. For instance, how would the populations be affected by a warming of +2 degrees? Such a prediction can be performed as shown below. 

```{r,results='asis'}
pred=prediction(X=data.frame(t=twoPopulations$t,
                             T1=twoPopulations$T1+2,
                             T2=twoPopulations$T2+2),
                spagFiles=c('P1_warming.spag','P2_warming.spag'),
                data.dir=workspace,doParametric=TRUE,doStructural=c(TRUE,TRUE))
BaM(workspace=workspace,mod=mod,data=data,pred=pred,doCalib=FALSE,doPred=TRUE)
```

The spaghetti resulting from this prediction can be plotted for P2 for instance (gray), and be compared with the population evolution associated with the original temperature series (red). 

```{r,results='asis'}
spag=utils::read.table(file.path(workspace,'P2_warming.spag'))
matplot(spag,type='l',col='gray')
lines(res$Y2_sim,col='red')
points(res$Y2_obs,col='red',pch=19)
```
